---
title: "file_corrections"
author: "Candi Pierce"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loads essential R packages for data manipulation (dplyr), reading files (readxl, readr), string operations (stringr), and custom SQL data access (utHelpR).
```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(utHelpR)
library(readxl)
library(readr)
library(stringr)
```

We received a correction file for the FVT/GE files submitted for 2022–2023 and 2023–2024. This script applies those corrections to the original files.

Reads the FVT-GE cohort CSV file and retrieves multiple datasets from SQL files, including student details, CIP codes, financial aid, and budget data.
```{r, warning=FALSE, message=FALSE}
fvt_ge_cohort_2223_df <- read_csv("/Volumes/Share/IE/Secure/FVT-GE/new files/003671_20222023_check.csv")
fvt_ge_cohort_2324_df <- read_csv("/Volumes/Share/IE/Secure/FVT-GE/new files/003671_20232024_check.csv")
correction_file_2223_df <- read_xlsx("/Volumes/Share/IE/Secure/FVT-GE/new files/correction_file_2223.xlsx")
correction_file_2324_df <- read_xlsx("/Volumes/Share/IE/Secure/FVT-GE/new files/correction_file_2324.xlsx")
```

This code merges the 2022–2023 cohort data with a correction file, flags records with error code 49 as invalid, and removes unnecessary columns.
```{r}
cohort_2223_df <- fvt_ge_cohort_2223_df %>%
  left_join(correction_file_2223_df, join_by("College Student ID","CIP Code")) %>% 

mutate(`Invalid Flag` = case_when(`Error Code` == 49 ~ "T",TRUE ~ as.character(`Invalid Flag`))) %>% 
  select(-c(`Record Type.y`,`Award Year.y`,`Error Code`,`Error Description`)) 
```

This code writes the cleaned and corrected 2022–2023 cohort data to a CSV file.
```{r}
write_csv(cohort_2223_df, "/Volumes/Share/IE/Secure/FVT-GE/Corrected Files/003671_20222023.csv")
```

This code merges the 2023–2024 cohort data with its correction file, flags error code 49 as invalid, and drops specified columns.
```{r}
cohort_2324_df <- fvt_ge_cohort_2324_df %>%
  left_join(correction_file_2324_df, join_by("College Student ID","CIP Code")) %>% 

mutate(`Invalid Flag` = case_when(`Error Code` == 49 ~ "T",TRUE ~ as.character(`Invalid Flag`))) %>% 
  select(-c(`Program Attendance Status During Award Year.y`,`Record Type.y`,`Error Code`)) 
```

This code writes the cleaned and corrected 2023–2024 cohort data to a CSV file.
```{r}
write_csv(cohort_2324_df, "/Volumes/Share/IE/Secure/FVT-GE/Corrected Files/003671_20232024.csv")
```


