---
title: "file_prep"
author: "Candi Pierce"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(utHelpR)
library(readxl)
library(readr)
library(stringr)
```

```{r, warning=FALSE, message=FALSE}
fvt_ge_cohort_2223_df <- read_csv("/Volumes/Share/IE/Secure/FVT-GE/2223Cohort_003671_AY20222023T.csv")
student_list <- utHelpR::get_data_from_sql_file('student_list.sql')
student_cip <- utHelpR::get_data_from_sql_file('student_cip.sql')
student_record <- utHelpR::get_data_from_sql_file('student_record.sql')
period_budget <- utHelpR::get_data_from_sql_file('period_budget.sql')
aid_year_budget <- utHelpR::get_data_from_sql_file('aid_year_budget.sql')
financial_aid <- utHelpR::get_data_from_sql_file('financial_aid.sql')
charges_assessed <- utHelpR::get_data_from_sql_file('charges_assessed.sql')
term_aid_year <- utHelpR::get_data_from_sql_file('term_to_aid_year.sql')
```

# Create student file with cip codes with terms
```{r}
cip_record_df <- student_record %>% 
  left_join(student_cip, join_by(pidm, cipcode == cip_code)) %>% 
  select(-c(cip_count, cip_sequence))
```

## FVT-GE File 
```{r}
fvt_ge_file_layout_df <- fvt_ge_cohort_2223_df %>% 
  janitor::clean_names() %>% 
  select(student_first_name, student_middle_name, student_last_name, cipcode = cip_code, published_length_of_program,  date_student_completed_or_withdrew_from_program, program_enrollment_begin_date, invalid_flag, studentid = college_student_id)
```


## TA - Private Loans
Total Amount Student Received in Private Education Loans During Student's Entire Enrollment in the Program by cip_code
```{r}
total_student_loan_df <- charges_assessed %>%
  filter(catcode == 'F1') %>%  # Financial Aid Loans
  inner_join(cip_record_df, by = "pidm") %>%
  filter(
    (is.na(derived_last_term_code) & termid >= first_term_code) |
    (!is.na(derived_last_term_code) & termid >= first_term_code & termid < derived_last_term_code)
  ) %>%
  group_by(studentid, pidm, cipcode) %>%
  summarise(total_loans_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>% 
  arrange(studentid, cipcode)
```


## TA - Institutional Debt
```{r}
total_institutional_debt_df <- charges_assessed %>%
  filter(typeind == "C", catcode != 'RFD') %>%
  inner_join(cip_record_df, by = "pidm") %>%
  filter(
    (is.na(derived_last_term_code) & termid >= first_term_code) |
    (!is.na(derived_last_term_code) & termid >= first_term_code & termid < derived_last_term_code)
  ) %>%
  group_by(studentid, pidm, cipcode) %>%
  summarise(total_debt_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  filter(total_debt_amount > 0) %>% 
  arrange(studentid, cipcode)
```



## TA - Total Amount of Tuition & Fees  
```{r}
total_tuition_fees_df <- charges_assessed %>%
  filter(catcode %in% c("TUI", "TU1", "TUD", "TUG", "TUN", "FE1", "FE2", "FEE")) %>%
  inner_join(cip_record_df, by = "pidm") %>%
  filter(
    (is.na(derived_last_term_code) & termid >= first_term_code) |
    (!is.na(derived_last_term_code) & termid >= first_term_code & termid < derived_last_term_code)
  ) %>%
  group_by(studentid, pidm, cipcode) %>%
  summarise(total_tuition_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  filter(total_tuition_amount > 0) %>%
  arrange(pidm, cipcode)
```



## TA - Total Amount of Allowance for Books and Supplies
```{r}
# Step 1: Filter for Books and Supplies
books_supplies <- aid_year_budget %>%
  filter(component == "B+S")

# Step 2: Join with term_aid_year to get termid for each aidyear
books_with_terms <- books_supplies %>%
  inner_join(term_aid_year, by = "aidyear")  # term_aid_year has columns: termid, aidyear

# Step 3: Join with cip_record_df by pidm only, then filter by term boundaries
books_split_df <- books_with_terms %>%
  inner_join(cip_record_df, by = "pidm") %>%
  filter(
    (is.na(derived_last_term_code) & termid >= first_term_code) |
    (!is.na(derived_last_term_code) & termid >= first_term_code & termid < derived_last_term_code)
  ) %>%
  mutate(
    split_amount = case_when(
      str_ends(termid, "30") ~ 0,
      str_ends(termid, "20|40") ~ aidamount / 2,
      TRUE ~ 0
    )
  )

# Step 4: Summarize totals by studentid, pidm, cipcode
total_books_supplies_df <- books_split_df %>%
  group_by(studentid, pidm, cipcode) %>%
  summarise(total_books_amount = sum(split_amount, na.rm = TRUE), .groups = "drop") %>%
  filter(total_books_amount > 0) %>%
  arrange(pidm, cipcode)
```



## TA - Total Amount of Grants & Scholarships Received
```{r}
total_grants_scholarships_df <- charges_assessed %>%
  filter(catcode %in% c("FA", "FA1", "FAN", "FAW", "FSW", "FWN", "FWR")) %>%
  inner_join(cip_record_df, by = "pidm") %>%
  filter(
    (is.na(derived_last_term_code) & termid >= first_term_code) |
    (!is.na(derived_last_term_code) & termid >= first_term_code & termid < derived_last_term_code)
  ) %>%
  group_by(studentid, pidm, cipcode) %>%
  summarise(total_grants_amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  filter(total_grants_amount > 0)%>%
  arrange(pidm, cipcode)
```




## Student file population with term and major
```{r}
cip_student_file_df <- cip_record_df %>% 
  select(pidm, studentid, cipcode) %>% 
  distinct() %>% 
  left_join(total_student_loan_df, by = c("pidm", "studentid", "cipcode")) %>% 
  left_join(total_institutional_debt_df, by = c("pidm", "studentid", "cipcode")) %>%
  left_join(total_books_supplies_df, by = c("pidm", "studentid", "cipcode")) %>%
  left_join(total_tuition_fees_df, by = c("pidm", "studentid", "cipcode")) %>%
  left_join(total_grants_scholarships_df, by = c("pidm", "studentid", "cipcode")) %>% 
  select(-pidm)
```




```{r}
final_TA_file_df <- fvt_ge_cohort_2223_df %>% 
  left_join(cip_student_file_df, by = c("College Student ID" = "studentid", "CIP Code" = "cipcode")) %>% 
  mutate(`Total Amount Student Received in Private Education Loans During Student's Entire Enrollment in the Program` =
           coalesce(total_loans_amount, 0)) %>% 
   mutate(`Total Amount of Institutional Debt During Student's Entire Enrollment in the Program` =
           coalesce(total_debt_amount, 0)) %>% 
  mutate(`Total Amount of Tuition & Fees Assessed During Student's Entire Enrollment in the Program` = coalesce(total_tuition_amount, 0)) %>% 
  
mutate(
    `Total Amount of Allowance for books, supplies, and equipment included in the student's title IV, HEA COA During Student's Entire Enrollment in the Program` =
      if_else(
        `Total Amount Student Received in Private Education Loans During Student's Entire Enrollment in the Program` == 0 &
        `Total Amount of Institutional Debt During Student's Entire Enrollment in the Program` == 0 &
        `Total Amount of Tuition & Fees Assessed During Student's Entire Enrollment in the Program` == 0,
        0,
        round(coalesce(total_books_amount, 0))
      )
  ) %>% 
  mutate(`Total Amount of Grants and Scholarships the student received During Student's Entire Enrollment in the Program` = coalesce(total_grants_amount, 0)) %>% 
  select(!c(total_loans_amount, total_debt_amount, total_tuition_amount, total_books_amount, total_grants_amount)) %>% 
  mutate(`Invalid Flag` = if_else(
      `Total Amount Student Received in Private Education Loans During Student's Entire Enrollment in the Program` == 0 &
      `Total Amount of Institutional Debt During Student's Entire Enrollment in the Program` == 0 &
      `Total Amount of Tuition & Fees Assessed During Student's Entire Enrollment in the Program` == 0 &
      `Total Amount of Allowance for books, supplies, and equipment included in the student's title IV, HEA COA During Student's Entire Enrollment in the Program` == 0 &
      `Total Amount of Grants and Scholarships the student received During Student's Entire Enrollment in the Program` == 0,
      "Yes", "No"
    )
  ) %>% 
mutate(
    `Comprehensive Transition and Postsecondary (CTP) Program Indicator` = 
      tidyr::replace_na(as.character(`Comprehensive Transition and Postsecondary (CTP) Program Indicator`), "No")) %>% 
    
    mutate(`Approved Prison Education Program Indicator` = 
      tidyr::replace_na(as.character(`Approved Prison Education Program Indicator`), "No")) %>% 

mutate(`Annual Cost of Attendance (COA)` = as.numeric(`Annual Cost of Attendance (COA)`)) %>%
  mutate(`Annual Cost of Attendance (COA)` = tidyr::replace_na(`Annual Cost of Attendance (COA)`, 0)) %>% 
  
  mutate(`Tuition and Fees Amount for Award Year being Reported` = as.numeric(`Tuition and Fees Amount for Award Year being Reported`)) %>%
  mutate(`Tuition and Fees Amount for Award Year being Reported` = tidyr::replace_na(`Tuition and Fees Amount for Award Year being Reported`, 0)) %>% 
  
  mutate(`Residency Tuition Status by State or District` = as.numeric(`Residency Tuition Status by State or District`)) %>%
  mutate(`Residency Tuition Status by State or District` = tidyr::replace_na(`Residency Tuition Status by State or District`, 0)) %>% 
  
  mutate(`Allowance for Books, Supplies, and Equipment` = as.numeric(`Allowance for Books, Supplies, and Equipment`)) %>%
  mutate(`Allowance for Books, Supplies, and Equipment` = tidyr::replace_na(`Allowance for Books, Supplies, and Equipment`, 0)) %>% 
  
  mutate(`Allowance for Housing and Food` = as.numeric(`Allowance for Housing and Food`)) %>%
  mutate(`Allowance for Housing and Food` = tidyr::replace_na(`Allowance for Housing and Food`, 0)) %>% 
  
  mutate(`Institutional Grants and Scholarships` = as.numeric(`Institutional Grants and Scholarships`)) %>%
  mutate(`Institutional Grants and Scholarships` = tidyr::replace_na(`Institutional Grants and Scholarships`, 0)) %>% 
  
  mutate(`Other State, Tribal, or Private Grants` = as.numeric(`Other State, Tribal, or Private Grants`)) %>%
  mutate(`Other State, Tribal, or Private Grants` = tidyr::replace_na(`Other State, Tribal, or Private Grants`, 0)) %>% 
  
  mutate(`Private Loans Amount` = as.numeric(`Private Loans Amount`)) %>%
  mutate(`Private Loans Amount` = tidyr::replace_na(`Private Loans Amount`, 0))

```








